\documentclass[11pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath, amssymb, amsthm}
\usepackage{geometry}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{enumitem}
\usepackage[table,xcdraw]{xcolor}
\usepackage{fancyhdr}

% Page Setup
\geometry{margin=1in}
\pagestyle{fancy}
\fancyhf{}
\lhead{CLL782: Process Optimization}
\rhead{Module 3.1: Environmental Load Modeling}
\cfoot{\thepage}

% Title Info
\title{\textbf{Process Optimization Project Report}\\
\large Module 3.1: Quantifying Environmental Load for Sustainable "Rendezvous"}
\author{Your Name \\ Entry Number: 202XXXXX \\ Department of Chemical Engineering, IIT Delhi}
\date{\today}

\begin{document}

\maketitle

% Declaration of Tool Usage
\section*{Declaration of Tool Usage}
I declare that in completing this assignment:
\begin{itemize}
    \item I used an LLM-based tool (Gemini) for assistance in:
    \begin{itemize}
        \item Brainstorming model formulations and exploring nonlinear crowding effects.
        \item Formatting the mathematical expressions and generating LaTeX code.
        \item Verifying the symbolic differentiation of the proposed objective function.
    \end{itemize}
    \item I understand the submitted solution fully.
    \item I can explain and justify every part of my code and reasoning.
    \item I have verified all results independently.
\end{itemize}

\vspace{0.5cm}
\hrule
\vspace{0.5cm}

\tableofcontents
\newpage

\section{Introduction}
The "Rendezvous" cultural festival at IIT Delhi presents a significant logistical and environmental challenge. As part of the Sustainability Task Force, our objective in Module 3.1 is to develop a robust mathematical model to quantify the total environmental load ($E$) of the festival. This model balances exploration of complex system dynamics (crowding, economies of scale) with the exploitation of rigorous optimization techniques to identify trade-offs and guide sustainable decision-making.

\section{Nomenclature}
The variables and parameters used in the mathematical model are defined in Table \ref{tab:nomenclature}.

\begin{table}[h]
\centering
\caption{Nomenclature Table}
\label{tab:nomenclature}
\begin{tabular}{@{}llcc@{}}
\toprule
\textbf{Symbol} & \textbf{Description} & \textbf{Units} & \textbf{Type} \\ \midrule
$N$ & Number of attendees & persons & Decision Var \\
$S$ & Number of food stalls/vendors & count & Decision Var \\
$A$ & Total event activities & activity-hours & Decision Var \\
$E$ & Total environmental load & kg CO$_2$-eq & Objective Fn \\
$\alpha_1$ & Base per-capita impact coefficient & kg CO$_2$-eq/person & Parameter \\
$\alpha_2$ & Base per-stall impact coefficient & kg CO$_2$-eq/stall & Parameter \\
$\gamma_{NS}$ & Congestion penalty coefficient & kg CO$_2$-eq $\cdot$ stall/person$^2$ & Parameter \\
$\beta_1$ & Nonlinear crowding exponent (attendees) & dimensionless & Parameter \\
$\beta_2$ & Nonlinear scaling exponent (stalls) & dimensionless & Parameter \\
$\epsilon$ & Grid emission factor & kg CO$_2$/kWh & Parameter \\
$\tau$ & Average event duration per attendee & hours & Parameter \\
$\xi$ & Sustainability Efficiency Metric & kg CO$_2$-eq/(person$\cdot$hr) & Derived Var \\
\bottomrule
\end{tabular}
\end{table}

\section{Assumptions and Justifications}
To develop a tractable yet expressive model, the following assumptions are made based on Life Cycle Assessment (LCA) principles and chemical engineering process economics (Letterman, 1980):

\begin{enumerate}
    \item \textbf{A1: Nonlinear Crowding Effects (Diseconomies of Scale)}. 
    \textit{Justification}: Environmental impact does not scale linearly with attendee density. As crowd density increases, waste collection efficiency drops, littering increases (overflowing bins), and resource competition leads to higher per-capita waste generation. We model this with a superlinear term $N^{\beta_1}$ where $\beta_1 > 1$.
    
    \item \textbf{A2: Economie of Scale in Infrastructure}.
    \textit{Justification}: High-fixed-cost infrastructure (generators, stages) becomes more efficient per unit of activity as scale increases. However, the total load still grows. For activities $A$, we use a sublinear power term $A^{0.8}$ for base energy load, reflecting standard equipment scaling laws (0.6-factor rule).
    
    \item \textbf{A3: Congestion-Induced Waste Leakage}.
    \textit{Justification}: When the ratio of attendees to service points (stalls $S$) is high ($N/S \gg \text{capacity}$), "leakage" occurs in the form of unsegregated waste and littering. This is modeled as a penalty term proportional to $N^2/S$.
    
    \item \textbf{A4: Sub-component Decomposition}.
    \textit{Justification}: The total load $E$ is the sum of Energy ($E_{energy}$), Waste ($W_{total}$), and Emissions ($E_{emissions}$), adhering to ISO 14040 LCA framework boundaries (Cradle-to-Grave for site operations).
\end{enumerate}

\section{Mathematical Model Formulation}

\subsection{Objective Function Construction}
We propose a multi-variable objective function $E(N, S, A)$ comprising three components: Linear Base Load ($E_{base}$), Nonlinear Scale Effects ($E_{scale}$), and Interaction/Congestion Penalties ($E_{inter}$).

\begin{equation}
    E(N, S, A) = E_{base} + E_{scale} + E_{inter}
\end{equation}

Substituting the terms:
\begin{equation} \label{eq:full_model}
    E(N, S, A) = (\alpha_1 N + \alpha_2 S + \alpha_3 A) + (\beta_N N^{1.3} + \beta_S S^{1.2} + \beta_A A^{0.8}) + \left( \gamma_{NS} \frac{N^2}{S} \right)
\end{equation}

Where:
\begin{itemize}
    \item $\alpha_1 N$: Direct resource consumption per attendee.
    \item $N^{1.3}$: The superlinear crowding term.
    \item $\gamma_{NS} \frac{N^2}{S}$: The \textbf{Congestion Penalty}. As $S \to 0$ for fixed $N$, environmental load explodes due to waste management failure. As $S \to \infty$, this term vanishes (perfect service), but the direct cost $\alpha_2 S$ increases. This creates a trade-off.
\end{itemize}

\subsection{Sub-Component Integration}
The aggregate function $E$ is derived from specific engineering estimates:

\begin{align}
    E_{energy} &= \epsilon \cdot \tau \cdot (P_{light} N + P_{stall} S + P_{stage} A) \\
    W_{total} &= \phi_{waste} \cdot N \cdot (1 + k_{crowd} \frac{N}{S}) \label{eq:waste_model} \\
    CO2e &= E_{transp} + E_{gen} + E_{mat}
\end{align}

Equation (\ref{eq:waste_model}) explicitly shows how waste generation rate depends on the crowding ratio $N/S$.

\section{Optimization Analysis}

\subsection{Unconstrained Optimization of $E$}
We seek to minimize $E(N, S, A)$ over the domain $N, S, A \ge 0$.

\textbf{Gradient Check}:
\begin{align}
    \frac{\partial E}{\partial N} &= \alpha_1 + 1.3 \beta_N N^{0.3} + \frac{2 \gamma_{NS} N}{S} > 0 \quad (\text{always positive for } N,S > 0) \\
    \frac{\partial E}{\partial A} &= \alpha_3 + 0.8 \beta_A A^{-0.2} > 0
\end{align}
Since the gradient components for productivity variables ($N, A$) are strictly positive, the unconstrained global minimum lies at the trivial solution:
\begin{equation}
    (N^*, S^*, A^*) = (0, 0, 0)
\end{equation}
\textbf{Insight}: Minimizing total environmental footprint to zero implies having no festival. This formulation, while chemically accurate, does not capture the \textit{value} of the event.

\subsection{Sub-Optimization: Optimal Stall Count}
However, for a \textbf{fixed number of attendees} $N$, there exists an optimal infrastructure level $S^*$. Differentiating $E$ with respect to $S$:

\begin{equation}
    \frac{\partial E}{\partial S} = \alpha_2 + 1.2 \beta_S S^{0.2} - \gamma_{NS} \frac{N^2}{S^2}
\end{equation}

Setting $\frac{\partial E}{\partial S} = 0$:
\begin{equation}
    \alpha_2 + 1.2 \beta_S S^{0.2} = \gamma_{NS} \frac{N^2}{S^2}
\end{equation}

The LHS represents the \textit{Marginal Environmental Cost} of adding a stall (embodied carbon, power). The RHS represents the \textit{Marginal Environmental Benefit} of reducing congestion (less litter). 
The unique intersection $S^*(N)$ defines the optimal resource allocation for any crowd size. 
Approximating for small nonlinearities ($\beta_S \approx 0$):
\begin{equation}
    S^* \approx N \sqrt{\frac{\gamma_{NS}}{\alpha_2}}
\end{equation}
This square-root law suggests that infrastructure should scale linearly with crowd size to maintain constant congestion levels $(N/S)$, but considering base costs, an optimal ratio emerges.

\subsection{Sustainability Efficiency (Part e)}
To address the trivial minimum problem, we refine the objective to minimize \textbf{Impact Intensity} or maximize \textbf{Sustainability Efficiency} ($\xi$):

\begin{equation}
    \text{minimize } \xi(N, S, A) = \frac{E(N, S, A)}{N \cdot \tau}
\end{equation}

This function represents the environmental cost per "attendee-hour".
\begin{align}
    \xi &= \frac{\alpha_1}{\tau} + \frac{\alpha_2 S}{N\tau} + \frac{\alpha_3 A}{N\tau} + \frac{\gamma_{NS} N}{S\tau} + \dots
\end{align}
Minimizing $\xi$ with respect to $N$ now involves trade-offs between fixed environmental costs (stalls, stages) which benefit from economies of scale (larger $N$), and crowding costs ($\gamma_{NS} N/S$) which penalize scale. This formulation yields a non-trivial measure of the "greenest" event scale.

\section{Preliminary Insights and Discussion}
\begin{itemize}
    \item \textbf{Crowding is Environmentally Expensive}: The term $\gamma_{NS} N^2/S$ is the dominant driver of inefficiency at large scales. Investing in sufficient infrastructure ($S$) is not just operationally necessary but environmentally critical to prevent waste leakage.
    \item \textbf{Economies of Scale}: The sublinear energy term for activities ($A^{0.8}$) suggests that consolidating entertainment into fewer, high-quality events is greener than many small, fragmented setups.
    \item \textbf{Critical Ratio}: The analysis reveals a critical Attendee-to-Stall ratio ($N/S)_{crit} = \sqrt{\alpha_2/\gamma_{NS}}$. Operating above this ratio causes environmental load to spike due to congestion.
\end{itemize}

\section{References}
\begin{enumerate}
    \item Edgar, T.F., Himmelblau, D.M., \& Lasdon, L.S. (2001). \textit{Optimization of Chemical Processes}, 2nd ed. McGraw-Hill.
    \item Letterman, R.D. (1980). Economic analysis of granular-bed filtration. \textit{Journal of the Environmental Engineering Division}, 106, 279–291.
    \item ISO 14040:2006. Environmental management — Life cycle assessment — Principles and framework.
    \item Prakash, O. (2025). \textit{Course Lecture Notes: Project Details}, CLL782/CHL7204, IIT Delhi.
\end{enumerate}

\end{document}
